Multiway Spherical Clustering via \\ Degree-Corrected Tensor Block Models

### 2022-01-18 tensor_clustering/others/ieee_v2.tex
organize

\begin{example}[Positivity of degree parameters]~\label{ex:positive}
Here we provide an example to show the positivity constraints on $\mtheta$ incurs no loss on the model flexibility. 
Consider an order-3 dTBM with core tensor $\tS = 1$ and degree $\mtheta = (1,1,-1, -1)^T$. We have the mean tensor 
\begin{equation}
    \tX = \tS \times_1 \mTheta \mM \times_2 \mTheta \mM \times_3 \mTheta \mM,
\end{equation}
where $\mTheta = \text{diag}(\mtheta)$ and $\mM = (1,1,1,1)^T$. Note that $\tX \in \bbR^{4 \times 4 \times 4}$ is a 1-block tensor with \emph{mixed-signed} degree $\mtheta$, and the mode-3 slices of $\tX$ are
\begin{equation}
    \tX_{::1} = \tX_{::2} = - \tX_{::3} = -\tX_{::4}  = \begin{bmatrix}
    1 & 1& -1& -1 \\
    1 & 1& -1& -1 \\
    -1& -1&1 & 1\\
     -1& -1&1 & 1
    \end{bmatrix}.
\end{equation}
Now, instead of original decomposition, we encode $\tX$ as a 2-block tensor with \emph{positive-signed} degree. Specifically, we write
\begin{equation}
     \tX = \tS' \times_1 \mTheta' \mM' \times_2 \mTheta' \mM'  \times_3 \mTheta' \mM' , 
\end{equation}
where $\mTheta' = \text{diag}(\mtheta') = \text{diag}(1,1, 1, 1)$, the core tensor $\tS' \in \bbR^{2 \times 2 \times 2}$ has mode-3 slices, and the membership matrix $\mM'\in\{0,1\}^{4\times 2}$ defines the clustering $z'\colon[4]\to[2]$,
\begin{equation}
    \tS'_{::1} = -\tS'_{::2} = \begin{bmatrix}
    1 & -1 \\
    -1 & 1
    \end{bmatrix}, \quad \mM' = \begin{bmatrix}
   1 & 0 \\
   1 &0 \\
    0 & 1 \\
    0&1
    \end{bmatrix}.
\end{equation}
 The triplet $(z',\tS', \mtheta')$ lies in our parameter space~\eqref{eq:family}. In general, we can always reparameterize a block-$r$ tensor with mixed-signed degree using a block-$2r$ tensor with positive-signed degree. Since we assume $r=\tO(1)$ throughout the paper, the splitting does not affect the error rates of our interest.\end{example}

@@

\begin{algorithm}[h!]
\caption*{\bf Algorithm: Multiway spherical clustering for degree-corrected tensor block model }
\vspace{.15cm}
\begin{algorithmic}[1]
\Algphase{Sub-algorithm 1: Weighted higher-order initialization}
\INPUT Observation $\tY \in \bbR^{p\times \cdots \times p}$, cluster number $r$, relaxation factor $\eta > 1$ in $k$-means clustering.
\State Compute factor matrix $ \mU_{\text{pre}} = \text{SVD}_{r} (\text{Mat}(\tY))$ and the $(K-1)$-mode projection $
\tX_{\text{pre}} = \tY \times_1   \mU_{\text{pre}} \mU_{\text{pre}}^T \times_2 \cdots \times_{K-1}  \mU_{\text{pre}} \mU_{\text{pre}}^T.$
\State Compute factor matrix $\hat \mU = \text{SVD}_{r}(\text{Mat}(\tX_{\text{pre}}))$ and denoised tensor
$\hat \tX = \tY \times_1 \hat \mU \hat \mU^T \times_2 \cdots \times_K \hat \mU \hat \mU^T.$
\State {Let $\hat \mX = \text{Mat}(\hat \tX)$ and $S_0=\{i \in [p]: \onormSize{}{\hat \mX_{i:}} = 0\}$. Set $\hat z(i)$ randomly in $[r]$ for $i \in S_0$.}
\State{For all $i\in S_0^c$, compute normalized rows
$\hat \mX_{i:}^s :=\onormSize{}{\hat \mX_{i:}}^{-1} \hat \mX_{i:}.$
}
\State {Solve the clustering $\hat z\colon [p]\to[r]$ and centroids $ (\hat \mx_j)_{j\in[r_k]}$ using weighted $k$-means, such that}
\begin{align}
    &\sum_{i \in S_0^c }  \onormSize{}{\hat \mX_{i:}}^2 \onormSize{}{\hat \mX_{i:}^s - \hat \mx_{\hat z(i)} }^2 
    \leq 
    \eta \min_{\substack{\bar \mx_j, j\in[r], \bar z(i),i\in S_0^{c}}} \sum_{i \in S^c } \onormSize{}{\hat \mX_{i:}}^2 \onormSize{}{ \hat \mX_{i:}^s -   \bar \mx_{\bar z(i)}}^2.
\end{align}
\OUTPUT Initial clustering $z^{(0)} \leftarrow \hat z$.

@@

\begin{thm}[Error for angle-based iteration]\label{thm:refinement} Consider the setup as in Theorem~\ref{thm:initial}. Assume the local linear stability of degree holds in the neighborhood $\tN(z,\epsilon)$ for $\epsilon = \log^{-1}p$. Suppose 
$r=\tO(1)$ and $\text{SNR} \geq \tilde C p^{-K/2}\log p$ for some large constant $\tilde C$. Let $z^{(t)}$ denote the $t$-th iteration output in Sub-algorithm~\hyperref[alg:main]{2} with initialization $z^{(0)}$ from Sub-algorithm~\hyperref[alg:main]{1}. With probability going to 1, there exists a contraction parameter $\rho \in (0,1)$ such that 
\begin{align}\label{eq:final}
    \ell(z, \hat z^{(t+1)}) \lesssim &\ \KeepStyleUnderBrace{
   \text{SNR}^{-1}
    \exp\of{- \frac{p^{K-1}\text{SNR}}{r^{K-1}}}}_{\substack{\text{statistical error}} }+ \KeepStyleUnderBrace{ \rho^t \ell(z, z^{(0)}). }_{\substack{\text{computational error}}}
\end{align}
\end{thm}
From the conclusion~\eqref{eq:final}, we find that the iteration error is decomposed into two parts: statistical error and computational error. The statistical error is unavoidable with noisy data regardless $t$, whereas the computational error decays in an exponential rate as the number of iterations $t \rightarrow \infty$.

@@

{\bf Extension for general dTBMs.} Our two-stage algorithm is able to be extended for the general (asymmetric) dTBMs. Specifically, in the Sub-algorithm~\hyperref[alg:main]{1}, we make the following changes: (1)
Replace the matrcization $\mat(\tY)$ by $\mat_k(\tY)$; (2) Repeat the Steps 1-5 with mode-specified cluster number $r_k$; (3) Obtain the collection initialization $ \{z^{(0)}_k\}_{k=1}^K$. In the Sub-algorithm~\hyperref[alg:main]{2}, we make the following changes: (1) Take the collection $\{z^{(0)}_k\}_{k=1}^K$ as input, and update the block tensor $\tS^{(t)}$ with the collection $\{ z_k^{(t)}\}_{k=1}^K$, $
        \tS^{(t)}(i_1,\ldots, i_K) =  \text{Ave} \{\tY(i_1,\ldots,i_K): z^{(t)}_k(i_k) = j_k, k \in [K]\}$;
(2) Calculate reduced tensor $\tY^{\text{d}}_k$ for each mode via
\begin{align}
    &\tY^{\text{d}}_k(a_1,\ldots,a_{k-1}, i ,a_{k+1},\ldots,a_K) 
    = \text{Ave}\{\tY(i_1,\ldots,i_{k-1},i,i_{k+1},\ldots,i_K): z^{(t)}(i_j) = a_j, j \neq k \}; 
\end{align}

@@

{\bf Hyper-parameter selection.} In our theoretical analysis, we have assumed the true cluster number $r$ is given to our algorithm. In practice, the cluster number $r$ is often unknown, and we now propose a method to choose $r$ from data. We impose the Bayesian information criterion (BIC) and choose the cluster number that minimizes BIC; i.e., under the symmetric Gaussian dTBM~\eqref{eq:model_tensor},
\begin{align}\label{eq:BIC}
 \hat r &= \argmin_{r \in \mathbb{Z}_+ } \of{p^K \log(\onormSize{}{\hat \tX - \tY}_F^2)  + p_e(r) K \log p},\\
 \text{where}  \quad \hat \tX &= \hat \tS(r) \times_1 \hat \mTheta(r) \hat \mM(r) \times_2 \cdots \times_K \hat \mTheta(r) \hat \mM(r),
\end{align}
where the triplet $(\hat z(r), \hat \mS(r),\hat \mtheta(r))$ are estimated parameters with cluster number $r$, and $p_e (r)= r^K + p(\log r + 1) - r$ is the effective number of parameters. Note that we have added the argument $(r)$ to related quantities as functions of $r$. In particular, the estimate $\hat \mtheta(r)$ in~\eqref{eq:BIC} is obtained by first calculating the reduced tensor $\hat \tY^{\text{d}}$ with $\hat z(r)$, and then normalizing the row norms $\onormSize{}{\hat \mY_{i:}^{\text{d}}}$ to 1 in each cluster; i.e., 
\begin{equation}
 \hat \mtheta(r)=(\hat \theta(1,r),\ldots,\hat \theta(p,r))^T,\quad\text{where}\quad   \hat \theta(i,r) = \frac{\onormSize{}{\hat \mY^{\text{d}}(r)_{i:}}}{\sum_{j: \hat z(j,r) =  \hat z(i,r)} \onormSize{}{\hat \mY^{\text{d}}(r)_{j:}}},
\end{equation}
where $\hat \mY^{\text{d}}(r) = \mat(\hat \tY^{\text{d}}(r))$, $\hat \tY^{\text{d}}(r)(i,a_2,\ldots,a_K) =  \text{Ave}\{\tY(i,i_2,\ldots,i_K): \hat z(i_k, r) = a_k, k \neq 1 \}$, and $\hat z(i,r)$ denotes the community label for the $i$-th node with given cluster number $r$. We evaluate the performance of the BIC criterion in Section~\ref{subsec:num_theory}.

@@

\begin{table}[hbt]
\centering
    \begin{tabularx}{\textwidth}{c *{8}{Y}}
    \toprule
    Settings & \multicolumn{2}{c}{$p = 50, \sigma^2 = 0.25$} & \multicolumn{2}{c}{$p = 50, \sigma^2 = 1$} & \multicolumn{2}{c}{$p = 80, \sigma^2 = 0.25$} & \multicolumn{2}{c}{$p = 80, \sigma^2 = 1$}\\
    \cmidrule(lr){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9}
         True cluster number $r$ & 2 & 4 & 2  & 4 & 2  & 4 & 2 & 4  \\
         \midrule
         Estimated cluster number $\hat r$ &  2(0)  & 3.9(0.25)& 2(0)    & 3.1(0.52) & 2(0)    & 4(0)   & 2(0)    & 3.9(0.31)   \\
     \bottomrule
    \end{tabularx}
    \caption{Estimated cluster number given by BIC criterion under the low noise level $(\sigma^2 = 0.25)$ and high noise level $(\sigma^2 = 0.5)$ settings. Numbers in parentheses are standard deviations of $\hat r$ over 30 replications.}
    \label{tab:select}
\end{table}

The third experiment evaluates the empirical performance of the BIC criterion to select unknown cluster number. We generate the data from an order-3 Gaussian model with $p = \{50,80\}$, $r = \{2,4\}$, and noise level $\sigma^2 \in \{ 0.25,1\}$. Table~\ref{tab:select} shows that our BIC criterion well chooses the true $r$ under most settings.  Note that the BIC slightly underestimates the true cluster number $(r = 4)$ with smaller dimension and higher noise $(p = 50, \sigma=1)$, and the accuracy immediately increases with larger dimension $p = 80$. The improvement follows from the fact that a larger dimension $p$ indicates a larger sample size in the tensor block model. Therefore, we conclude that BIC criterion is a reasonable way to tune the cluster number.

@@

\subsection{Proof sketch of Theorem 5}
The proof of Theorem~\ref{thm:refinement} is inspired by the proof idea of \citet[Theorem 2]{han2020exact}. We develop extra polar-coordinate based techniques with angle gap characterization to address the nuisance degree heterogeneity. We introduce an intermediate quantity called misclustering loss
\begin{equation}
     L^{(t)} = \frac{1}{p}  \sum_{i \in [p]} \theta(i) \sum_{b \in [r]}  \ind \offf{ z^{(t)}(i) = b } \onorm{ \off{ \mS_{ z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2,
\end{equation}
where the superscript $\cdot^{s}$ denotes the normalized vector; i.e., $\ma^s:=\ma/\onorm{\ma}$ if $\ma\neq 0$ and $\ma^{s}=0$ if $\ma=0$ for any vector $\ma$. We show that $L^{(t)}$ provides an upper bound for the misclassification error of interest via the inequality $\ell^{(t)}\leq {L^{(t)}\over \Delta^2_{\min}}$. Therefore, it suffices to control $L^{(t)}$. Further, we introduce the oracle estimators for core tensor under the true cluster assignment via 
\begin{equation}
    \tilde \tS = \tY \times_1 \mW^T \times_2 \cdots \times_K \mW^T, 
\end{equation}
where $\mW = \mM \of{ \text{diag}(\mone_{p}^T \mM) }^{-1}$ is the weighted true membership matrix. Let $ \mV = \mW^{\otimes (K-1)}$ denote the Kronecker product of $(K-1)$ copies of $\mW$ matrices, and we define the $t$-th iteration quantities $\mW^{(t)}, \mV^{(t)}$ corresponding to $\mM^{(t)}$ (or equivalently $z^{(t)}$). To evaluate $L^{(t+1)}$, we prove the bound
\begin{equation}\label{eq:proof_5_event}
    \ind \offf{ z^{(t+1)}(i) = b } = \ind \offf{       \onormSize{}{ [ \mY_{i:} \mV^{(t)}  ]^s - [\mS_{b:}^{(t)}]^s }^2 \leq \onormSize{}{ [ \mY_{i:} \mV^{(t)}  ]^s - [\mS_{z(i):}^{(t)}]^s }^2} \leq A_{ib} + B_{ib},
\end{equation}
where $\mY = \mat(\tY)$, $ \mS = \mat(\tS)$, $\mS^{(t)} = \mat(\tS^{(t)})$ and
\begin{align}
        A_{ib} &= \ind \offf{\ang{ \mE_{i:} \mV, \off{  \tilde \mS_{z(i):} }^s - \off{  \tilde \mS_{b:} }^s } \lesssim -  \onorm{ \off{ \mS_{z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2 },\\
        B_{ib} &= \ind \offf{\onorm{ \off{ \mS_{z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2 \lesssim F_{ib}^{(t)} + G_{ib}^{(t)} + H_{ib}^{(t)} }.
\end{align}
The terms $F_{ib}^{(t)}, G_{ib}^{(t)}, H_{ib}^{(t)}$ are controlled by $z^{(t)}, \tS^{(t)}$; see the detailed definitions in \eqref{eq:f}, \eqref{eq:g}, \eqref{eq:h}. Note that the event $A_{ib}$ only involves the oracle estimator independent of $t$, while all the terms related to the $t$-th iteration are in $B_{ib}$. Thus, the inequality~\eqref{eq:proof_5_event} decomposes the misclustering loss in the $(t+1)$-th iteration into the oracle loss and the loss in $t$-th iteration. This decomposition leads to the separation of statistical error and computational error in the final upper bound of Theorem~\ref{thm:refinement}.

Specifically, we prove the contraction inequality
\begin{equation}\label{eq:proof_5_ineq}
    L^{(t+1)} \leq M \xi + \rho L^{(t)}, \quad \xi = \frac{1}{p}  \sum_{i \in [p]} \theta(i) \sum_{b \in [r]}  A_{ib} \onorm{ \off{ \mS_{ z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2,
\end{equation}
where $M$ is a positive constant, $\rho \in (0,1)$ is the contraction parameter, and we call $\xi$ the oracle loss. Controlling the probability of event $B_{ib}$ and obtaining the $\rho L^{(t)}$ term in the right hand side of~\eqref{eq:proof_5_ineq} are the most challenging parts in the proof of Theorem~\ref{thm:refinement}. Note that the true and estimated core tensors are involved via their normalized rows such as $\mS_{a:}^s, \tilde \mS_{a:}^s, [\mS^{(t)}_{a:}]^s$. The Cartesian coordinate based analysis in \cite{han2020exact} is no longer applicable in our case. Instead, we use the polar-coordinate based analysis and the geometry property of trigonometric functions to derive the high probability upper bounds for $F_{ib}^{(t)}, G_{ib}^{(t)}, H_{ib}^{(t)}$.

@@

 If $\mS_{1,1:}$ is a zero vector, construct $\mtheta_1'$ such that $\mtheta'_{1,z_1^{-1}(1)} \neq \mtheta_{1,z_1^{-1}(1)}$. Let $\{z'_k\} = \{z_k\}$, $\tS' = \tS$, and $\mtheta'_k = \mtheta_k$ for all $k = 2, \ldots, K$. Then the triplet $\of{ \{z'_k\}, \tS', \{ \mtheta'_k \} }$ is distinct from $\of{ \{z_k\}, \tS, \{ \mtheta_k \} }$ up to label permutation. Similar conclusion holds when $\mS_{1,2:}$ is a zero vector.

@@

\begin{equation}\label{eq:cond_oracle}
    \xi \leq \exp\of{ - M \frac{\Delta_{\min}^2 p^{K-1}}{r^{K-1}}},
\end{equation}

>>>>

\begin{example}[Positivity of degree parameters]~\label{ex:positive}
Here we provide an example to show the positivity constraints on $\mtheta$ incurs no loss on the model flexibility. 
Consider an order-3 dTBM with core tensor $\tS = 1$ and degree $\mtheta = (1,1,-1, -1)^T$. We have the mean tensor 
\begin{equation}
    \tX = \tS \times_1 \mTheta \mM \times_2 \mTheta \mM \times_3 \mTheta \mM,
\end{equation}
where $\mTheta = \text{diag}(\mtheta)$ and $\mM = (1,1,1,1)^T$. Note that $\tX \in \bbR^{4 \times 4 \times 4}$ is a 1-block tensor with \emph{mixed-signed} degree $\mtheta$, and the mode-3 slices of $\tX$ are
\begin{equation}
    \tX_{::1} = \tX_{::2} = - \tX_{::3} = -\tX_{::4}  = \begin{bmatrix}
    1 & 1& -1& -1 \\
    1 & 1& -1& -1 \\
    -1& -1&1 & 1\\
     -1& -1&1 & 1
    \end{bmatrix}.
\end{equation}
Now, instead of original decomposition, we encode $\tX$ as a 2-block tensor with \emph{positive-signed} degree. Specifically, we write
\begin{equation}
     \tX = \tS' \times_1 \mTheta' \mM' \times_2 \mTheta' \mM'  \times_3 \mTheta' \mM' , 
\end{equation}
where $\mTheta' = \text{diag}(\mtheta') = \text{diag}(1,1, 1, 1)$, the core tensor $\tS' \in \bbR^{2 \times 2 \times 2}$ has mode-3 slices, and the membership matrix $\mM'\in\{0,1\}^{2\times 4}$ defines the clustering $z'\colon[4]\to[2]$,
\begin{equation}
    \tS'_{::1} = -\tS'_{::2} = \begin{bmatrix}
    1 & -1 \\
    -1 & 1
    \end{bmatrix}, \quad \mM' = \begin{bmatrix}
   1& 1 & 0 & 0\\
    0 &0 &  1& 1 
    \end{bmatrix}.
\end{equation}
 The triplet $(z',\tS', \mtheta')$ lies in our parameter space~\eqref{eq:family}. In general, we can always reparameterize a block-$r$ tensor with mixed-signed degree using a block-$2r$ tensor with positive-signed degree. Since we assume $r=\tO(1)$ throughout the paper, the splitting does not affect the error rates of our interest.\end{example}

@@

\begin{algorithm}[h!]
\caption*{\bf Algorithm: Multiway spherical clustering for degree-corrected tensor block model }
\vspace{.15cm}
\begin{algorithmic}[1]
\Algphase{Sub-algorithm 1: Weighted higher-order initialization}
\INPUT Observation $\tY \in \bbR^{p\times \cdots \times p}$, number of clusters $r$, relaxation factor $\eta > 1$ in $k$-means clustering.
\State Compute factor matrix $ \mU_{\text{pre}} = \text{SVD}_{r} (\text{Mat}(\tY))$ and the $(K-1)$-mode projection $
\tX_{\text{pre}} = \tY \times_1   \mU_{\text{pre}} \mU_{\text{pre}}^T \times_2 \cdots \times_{K-1}  \mU_{\text{pre}} \mU_{\text{pre}}^T.$
\State Compute factor matrix $\hat \mU = \text{SVD}_{r}(\text{Mat}(\tX_{\text{pre}}))$ and denoised tensor
$\hat \tX = \tY \times_1 \hat \mU \hat \mU^T \times_2 \cdots \times_K \hat \mU \hat \mU^T.$
\State {Let $\hat \mX = \text{Mat}(\hat \tX)$ and $S_0=\{i \in [p]: \onormSize{}{\hat \mX_{i:}} = 0\}$. Set $\hat z(i)$ randomly in $[r]$ for $i \in S_0$.}
\State{For all $i\in S_0^c$, compute normalized rows
$\hat \mX_{i:}^s :=\onormSize{}{\hat \mX_{i:}}^{-1} \hat \mX_{i:}.$
}
\State {Solve the clustering $\hat z\colon [p]\to[r]$ and centroids $ (\hat \mx_j)_{j\in[r_k]}$ using weighted $k$-means, such that}
\begin{align}
    &\sum_{i \in S_0^c }  \onormSize{}{\hat \mX_{i:}}^2 \onormSize{}{\hat \mX_{i:}^s - \hat \mx_{\hat z(i)} }^2 
    \leq 
    \eta \min_{\substack{\bar \mx_j, j\in[r], \bar z(i),i\in S_0^{c}}} \sum_{i \in S^c } \onormSize{}{\hat \mX_{i:}}^2 \onormSize{}{ \hat \mX_{i:}^s -   \bar \mx_{\bar z(i)}}^2.
\end{align}
\OUTPUT Initial clustering $z^{(0)} \leftarrow \hat z$.

@@

\begin{thm}[Error for angle-based iteration]\label{thm:refinement} Consider the setup as in Theorem~\ref{thm:initial}. Assume the local linear stability of degree holds in the neighborhood $\tN(z,\epsilon)$ for $\epsilon \geq \log^{-1}p$. Suppose 
$r=\tO(1)$ and $\text{SNR} \gtrsim  p^{-K/2}\log p$. Let $z^{(t)}$ denote the $t$-th iteration output in Sub-algorithm~\hyperref[alg:main]{2} with initialization $z^{(0)}$ from Sub-algorithm~\hyperref[alg:main]{1}. With probability going to 1, there exists a contraction parameter $\rho \in (0,1)$ such that 
\begin{align}\label{eq:final}
    \ell(z, \hat z^{(t+1)}) \lesssim &\ \KeepStyleUnderBrace{
   \text{SNR}^{-1}
    \exp\of{- \frac{p^{K-1}\text{SNR}}{r^{K-1}}}}_{\substack{\text{statistical error}} }+ \KeepStyleUnderBrace{ \rho^t \ell(z, z^{(0)}). }_{\substack{\text{computational error}}}
\end{align}
\end{thm}
From the conclusion~\eqref{eq:final}, we find that the iteration error is decomposed into two parts: statistical error and computational error. The statistical error is unavoidable with noisy data regardless $t$, whereas the computational error decays in an exponential rate as the number of iterations $t \rightarrow \infty$.

@@

{\bf Extension for general dTBMs.} Our two-stage algorithm is able to be extended for the general (asymmetric) dTBMs. Specifically, in the Sub-algorithm~\hyperref[alg:main]{1}, we make the following changes: (1)
Replace the matrcization $\mat(\tY)$ by $\mat_k(\tY)$; (2) Repeat the Steps 1-5 with mode-specified number of clusters $r_k$; (3) Obtain the collection initialization $ \{z^{(0)}_k\}_{k=1}^K$. In the Sub-algorithm~\hyperref[alg:main]{2}, we make the following changes: (1) Take the collection $\{z^{(0)}_k\}_{k=1}^K$ as input, and update the block tensor $\tS^{(t)}$ with the collection $\{ z_k^{(t)}\}_{k=1}^K$, $
        \tS^{(t)}(i_1,\ldots, i_K) =  \text{Ave} \{\tY(i_1,\ldots,i_K): z^{(t)}_k(i_k) = j_k, k \in [K]\}$;
(2) Calculate reduced tensor $\tY^{\text{d}}_k$ for each mode via
\begin{align}
    &\tY^{\text{d}}_k(a_1,\ldots,a_{k-1}, i ,a_{k+1},\ldots,a_K) 
    = \text{Ave}\{\tY(i_1,\ldots,i_{k-1},i,i_{k+1},\ldots,i_K): z^{(t)}(i_j) = a_j, j \neq k \}; 
\end{align}

@@

{\bf Hyper-parameter selection.} In our theoretical analysis, we have assumed the true number of clusters $r$ is given to our algorithm. In practice, the number of clusters $r$ is often unknown, and we now propose a method to choose $r$ from data. We impose the Bayesian information criterion (BIC) and choose the cluster number that minimizes BIC; i.e., under the symmetric Gaussian dTBM~\eqref{eq:model_tensor},
\begin{align}\label{eq:BIC}
 \hat r &= \argmin_{r \in \mathbb{Z}_+ } \of{p^K \log(\onormSize{}{\hat \tX - \tY}_F^2)  + p_e(r) K \log p},\\
 \text{where}  \quad \hat \tX &= \hat \tS(r) \times_1 \hat \mTheta(r) \hat \mM(r) \times_2 \cdots \times_K \hat \mTheta(r) \hat \mM(r),
\end{align}
where the triplet $(\hat z(r), \hat \mS(r),\hat \mtheta(r))$ are estimated parameters with cluster number $r$, and $p_e (r)= r^K + p(\log r + 1) - r$ is the effective number of parameters. Note that we have added the argument $(r)$ to related quantities as functions of $r$. In particular, the estimate $\hat \mtheta(r)$ in~\eqref{eq:BIC} is obtained by first calculating the reduced tensor $\hat \tY^{\text{d}}$ with $\hat z(r)$, and then normalizing the row norms $\onormSize{}{\hat \mY_{i:}^{\text{d}}}$ to 1 in each cluster; i.e., 
\begin{equation}
 \hat \mtheta(r)=(\hat \theta(1,r),\ldots,\hat \theta(p,r))^T,\quad\text{where}\quad   \hat \theta(i,r) = \frac{\onormSize{}{\hat \mY^{\text{d}}(r)_{i:}}}{\sum_{j: \hat z(j,r) =  \hat z(i,r)} \onormSize{}{\hat \mY^{\text{d}}(r)_{j:}}},
\end{equation}
where $\hat \mY^{\text{d}}(r) = \mat(\hat \tY^{\text{d}}(r))$, $\hat \tY^{\text{d}}(r)(i,a_2,\ldots,a_K) =  \text{Ave}\{\tY(i,i_2,\ldots,i_K): \hat z(i_k, r) = a_k, k \neq 1 \}$, and $\hat z(i,r)$ denotes the community label for the $i$-th node with given cluster number $r$. We evaluate the performance of the BIC criterion in Section~\ref{subsec:num_theory}.

@@

\begin{table}[hbt]
\centering
    \begin{tabularx}{\textwidth}{c *{8}{Y}}
    \toprule
    Settings & \multicolumn{2}{c}{$p = 50, \sigma^2 = 0.25$} & \multicolumn{2}{c}{$p = 50, \sigma^2 = 1$} & \multicolumn{2}{c}{$p = 80, \sigma^2 = 0.25$} & \multicolumn{2}{c}{$p = 80, \sigma^2 = 1$}\\
    \cmidrule(lr){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9}
         True number of clusters $r$ & 2 & 4 & 2  & 4 & 2  & 4 & 2 & 4  \\
         \midrule
         Estimated number of clusters $\hat r$ &  2(0)  & 3.9(0.25)& 2(0)    & 3.1(0.52) & 2(0)    & 4(0)   & 2(0)    & 3.9(0.31)   \\
     \bottomrule
    \end{tabularx}
    \caption{Estimated number of clusters given by BIC criterion under the low noise $(\sigma^2 = 0.25)$ and high noise $(\sigma^2 = 0.5)$ settings. Numbers in parentheses are standard deviations of $\hat r$ over 30 replications.}
    \label{tab:select}
\end{table}

The third experiment evaluates the empirical performance of the BIC criterion to select unknown number of clusters. We generate the data from an order-3 Gaussian model with $p = \{50,80\}$, $r = \{2,4\}$, and noise level $\sigma^2 \in \{ 0.25,1\}$. Table~\ref{tab:select} shows that our BIC criterion well chooses the true $r$ under most settings.  Note that the BIC slightly underestimates the true number of clusters $(r = 4)$ with smaller dimension and higher noise $(p = 50, \sigma=1)$, and the accuracy immediately increases with larger dimension $p = 80$. The improvement follows from the fact that a larger dimension $p$ indicates a larger sample size in the tensor block model. Therefore, we conclude that BIC criterion is a reasonable way to tune the number of clusters.

@@

\subsection{Proof sketch of Theorem 5}
The proof of Theorem~\ref{thm:refinement} is inspired by the proof idea of \citet[Theorem 2]{han2020exact}. We develop extra polar-coordinate based techniques with angle gap characterization to address the nuisance degree heterogeneity. We introduce an intermediate quantity called misclustering loss
\begin{equation}
     L^{(t)} = \frac{1}{p}  \sum_{i \in [p]} \theta(i) \sum_{b \in [r]}  \ind \offf{ z^{(t)}(i) = b } \onorm{ \off{ \mS_{ z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2,
\end{equation}
where the superscript $\cdot^{s}$ denotes the normalized vector; i.e., $\ma^s:=\ma/\onorm{\ma}$ if $\ma\neq 0$ and $\ma^{s}=0$ if $\ma=0$ for any vector $\ma$. We show that $L^{(t)}$ provides an upper bound for the misclassification error of interest via the inequality $\ell^{(t)}\lesssim {L^{(t)}\over \Delta^2_{\min}}$. Therefore, it suffices to control $L^{(t)}$. Further, we introduce the oracle estimators for core tensor under the true cluster assignment via 
\begin{equation}
    \tilde \tS = \tY \times_1 \mW^T \times_2 \cdots \times_K \mW^T, 
\end{equation}
where $\mW = \mM \of{ \text{diag}(\mone_{p}^T \mM) }^{-1}$ is the weighted true membership matrix. Let $ \mV = \mW^{\otimes (K-1)}$ denote the Kronecker product of $(K-1)$ copies of $\mW$ matrices, and we define the $t$-th iteration quantities $\mW^{(t)}, \mV^{(t)}$ corresponding to $\mM^{(t)}$ (or equivalently $z^{(t)}$). To evaluate $L^{(t+1)}$, we prove the bound
\begin{equation}\label{eq:proof_5_event}
    \ind \offf{ z^{(t+1)}(i) = b } = \ind \offf{       \onormSize{}{ [ \mY_{i:} \mV^{(t)}  ]^s - [\mS_{b:}^{(t)}]^s }^2 \leq \onormSize{}{ [ \mY_{i:} \mV^{(t)}  ]^s - [\mS_{z(i):}^{(t)}]^s }^2} \leq A_{ib} + B_{ib},
\end{equation}
where $\mY = \mat(\tY)$, $ \mS = \mat(\tS)$, $\mS^{(t)} = \mat(\tS^{(t)})$ and
\begin{align}
        A_{ib} &= \ind \offf{\ang{ \mE_{i:} \mV, \off{  \tilde \mS_{z(i):} }^s - \off{  \tilde \mS_{b:} }^s } \lesssim -  \onorm{ \off{ \mS_{z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2 },\\
        B_{ib} &= \ind \offf{\onorm{ \off{ \mS_{z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2 \lesssim F_{ib}^{(t)} + G_{ib}^{(t)} + H_{ib}^{(t)} }.
\end{align}
The terms $F_{ib}^{(t)}, G_{ib}^{(t)}, H_{ib}^{(t)}$ are controlled by $z^{(t)}, \tS^{(t)}$; see the detailed definitions in \eqref{eq:f}, \eqref{eq:g}, \eqref{eq:h}. Note that the event $A_{ib}$ only involves the oracle estimator independent of $t$, while all the terms related to the $t$-th iteration are in $B_{ib}$. Thus, the inequality~\eqref{eq:proof_5_event} decomposes the misclustering loss in the $(t+1)$-th iteration into the oracle loss and the loss in $t$-th iteration. This decomposition leads to the separation of statistical error and computational error in the final upper bound of Theorem~\ref{thm:refinement}.

Specifically, we prove the contraction inequality
\begin{equation}\label{eq:proof_5_ineq}
    L^{(t+1)} \lesssim \xi + \rho L^{(t)}, \quad \xi = \frac{1}{p}  \sum_{i \in [p]} \theta(i) \sum_{b \in [r]}  A_{ib} \onorm{ \off{ \mS_{ z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2,
\end{equation}
where $\rho \in (0,1)$ is the contraction parameter, and we call $\xi$ the oracle loss. Controlling the probability of event $B_{ib}$ and obtaining the $\rho L^{(t)}$ term in the right hand side of~\eqref{eq:proof_5_ineq} are the most challenging parts in the proof of Theorem~\ref{thm:refinement}. Note that the true and estimated core tensors are involved via their normalized rows such as $\mS_{a:}^s, \tilde \mS_{a:}^s, [\mS^{(t)}_{a:}]^s$. The Cartesian coordinate based analysis in \cite{han2020exact} is no longer applicable in our case. Instead, we use the polar-coordinate based analysis and the geometry property of trigonometric functions to derive the high probability upper bounds for $F_{ib}^{(t)}, G_{ib}^{(t)}, H_{ib}^{(t)}$.

@@

 If $\mS_{1,1:}$ is a zero vector, construct $\mtheta_1'$ such that $\mtheta'_{1,z_1^{-1}(1)} \neq \mtheta_{1,z_1^{-1}(1)}$. Let $\{z'_k\} = \{z_k\}$, $\tS' = \tS$, and $\mtheta'_k = \mtheta_k$ for all $k = 2, \ldots, K$. Then the triplet $\of{ \{z'_k\}, \tS', \{ \mtheta'_k \} }$ is distinct from \of{ \{z_k\}, \tS, \{ \mtheta_k \} }$ up to label permutation. Similar conclusion holds when $\mS_{1,2:}$ is a zero vector.

@@

\begin{equation}\label{eq:cond_oracle}
    \xi \lesssim \exp\of{ - \frac{\Delta_{\min}^2 p^{K-1}}{r^{K-1}}},
\end{equation}

\begin{equation}\label{eq:cond_intial}
    L^{(t)} \lesssim \frac{\Delta_{\min}^2}{r \log p}, \quad \text{for} \quad t = 0, 1, \ldots, T.
\end{equation}

Further, inequality~\eqref{eq:cond1} holds by replacing $\mV$ to $\mV^{(t)}$ and $\mW_{:a}$ to $\mW_{:a}^{(t),T}$ when initialization condition~\eqref{eq:cond_intial} holds.
\end{condition}
